---
title: "m&m"
author: "Ilaria Pretelli"
date: "4/3/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(rethinking)
library(rlist)
source("../1_simulation/1_simulation.R")
real_data <- list.load("../2_data_preparation/processed_data.RData")
trapcol <- "#5ca81e"
shellcol <-  "#ea5914"
seq_trait <- seq(0,3,0.01)

```


```{r setup variables}
#####################
nsamp <-100
#PRIOR PREDICTIVE SIMULATION
#Set trait values for age and possible times
AGE <- seq(0,3, 0.1) #trait
L <- c(0, 0.1, 1, 3) #trait
#create matrixes to save individual and trip effects
phi <- matrix(NA, nsamp, length(AGE))
psi <- rep(NA, nsamp)
#SIMULATED PRIORS
#general intercept
alpha <- rlnorm(nsamp, 0, 0.1)#prior
#trip
lambda <-  rexp(nsamp,1)
#individual_age
beta_a <- rlnorm(nsamp, 0, 1)#prior
gamma_a <- rlnorm(nsamp, 0, 1)#prior
#sigma lognormal
sigma <- rexp(nsamp, 1)
#ADD KNOWLEDGE
beta_ak <- rlnorm(nsamp, 0, 1)#prior
beta_k <- rnorm(nsamp, 0, 1)#prior
K <- matrix(NA, nsamp, length(AGE)) 
for(i in 1:nsamp){
  K[i,] <- beta_ak[i]*AGE #trait
}
beta_ab <- rlnorm(nsamp, 0, 1)#prior
beta_b <- rnorm(nsamp, 0, 1)#prior
B <- matrix(NA, nsamp, length(AGE)) 
for(i in 1:nsamp){
  B[i,] <- beta_ab[i]*AGE #trait
}

```


# SET UP MODEL

Foraging returns have been modeled as hurdle models, modeling separately (i) the probability of success (i.e. non zero returns) and (ii) the amount of returns (e.g. grams or calories collected) for a certain trip. In this case, the two types of foraging analyzed can be preliminary evaluated with just one part of a hurdle model. This is because, for hunting, we mainly model the probability that each trap has of collecting anything at all, i.e. the first part of a hurdle model is sufficient (at least for now). For collecting seashells, the second part of the hurdle model is enough, as when children set out for this activity, they ALWAYS return something--hence no need to model whether they'll capture anything at all.

Mathematically, whether a trap captures something between being checked one day and the following check is an outcome Bernoully distributed with probability $p$. The amount of shells collected by children, instead, is a Lognormally distributed quantity, where the mean of the distribution is $\mu$ and standard deviation is $\sigma$. 

$$ 
S \sim Bernoulli (p)\\
R \sim Lognormal ( \mu, \sigma)\\
$$

Now, $p$ and $\mu$ are composed of parallel parts, for clarity and comparability. Although combined differently and appropriately scaled, both of them are a combination of a general intercept, $\alpha$, an equation including the effects of individual level traits, designed by $\phi$, and an equation for trip level traits, designed by $\psi$.

$$
p = 1 - exp ( - \alpha * \phi_i * \psi_t  )\\
\mu = log(\alpha * \phi_i * \psi_t) \\
$$

Before understanding how these parts are tied together, let's delve into the individual and, then, trip level equations $\phi$ and $\psi$.

$\phi$ includes an individual level random effect, $\iota$, and a series of factors, which, multiplied, define the importance of one or more between age, knowledge and somatic traits. These factors are all built in the same way, where the maximum effect of each of these is equal to 1 and their effect is multiplied.  

The equation that describes the effect of age (equivalent to that describing the effect of knowledge and somatic traits) levels off at 1, and has a curved shape where the inclination of the curve is given by $\beta$ and how fast the curve sets off from the x axis is given by $\gamma$.
For a model with age effect only, $\phi$ is then equal to: 

$$
\phi = \iota_i * (1-exp(-\beta * age_i  )) ^ {\gamma}
$$
The following plot shows the effect on the curve of changing values of $\beta$ and $\gamma$ (respectively noted as b and g in the plot).
```{r shape of age effects}
#plot prior predictive simulation
plot(NULL, xlim = c(0,3), ylim = c (0,1), 
     xlab = "Age", ylab = "Proportion improvement")
#calculate per sample
for(i in 1:nsamp){
  phi[i,] <- (1-exp(-beta_a[i] * AGE  )) ^ gamma_a[i]
  lines( AGE,  phi[i,], col = col.alpha("cornflowerblue", 0.3))
}
lines (AGE, (1-exp(-1 * AGE  )) ^ 1,
       col = col.alpha("darkblue", 0.7))
text(1.5, 0.8,"b=1,g=1")
lines (AGE, (1-exp(-1 * AGE  )) ^ 0.1,
       col = col.alpha("darkblue", 0.7))
text(0.4, 0.9,"b=1,g=0.1")
lines (AGE, (1-exp(-1 * AGE  )) ^ 10,
       col = col.alpha("darkblue", 0.7))
text(2.5, 0.4,"b=1,g=10")
lines (AGE, (1-exp(-0.1 * AGE  )) ^ 1,
       col = col.alpha("darkblue", 0.7))
text(2.5, 0.2,"b=0.1,g=1")
lines (AGE, (1-exp(-3 * AGE  )) ^ 1,
       col = col.alpha("darkblue", 0.7))
text(0.4, 0.7,"b=3,g=1")
```

We can then include also the effect of knowledge and/or somatic traits. Each of these is scaled by an elasticity parameter ($\zeta$ and $\eta$ respectively ) and multiplied to the function shaping the effect of age. 

$$
\phi = \iota_i \ (1-exp(-\beta \ age_i  )) ^ {\gamma}\ knowledge_i\, ^ {\zeta} \ body_i\, ^ {\eta} \\
$$

Then $\phi$ is shaped by the effect of age, and scaled to include the effects of knowledge and body. $\zeta$ and $\eta$ can take any value, with positive numbers giving a diminishing return pattern to the trait in question and negative ones giving a pattern of increasing returns. 

```{r knowledge and body effects}
#plot prior predictive simulation
plot(NULL, xlim = c(0,3), ylim = c (0,3), 
     xlab = "Age", ylab = "Proportion improvement")
#calculate per sample
for(i in 1:nsamp){
  phi[i,] <- (1-exp(-beta_a[i] * AGE  )) ^ gamma_a[i] *
     mean(K) ^ beta_k[i]
  lines( AGE,  phi[i,], col = col.alpha("cornflowerblue", 0.3))
}

lines (AGE, (1-exp(-1 * AGE  )) ^ 1 ,
       col = col.alpha("darkblue", 0.7))
text(1.5, 0.8,"age only")
lines (AGE, (1-exp(-1 * AGE  )) ^ 1 * 
          mean(K) ^ 1,
       col = col.alpha("darkblue", 0.7))
text(1, 1.4,"with knowledge")
lines (AGE, (1-exp(-1 * AGE  )) ^ 1 * 
         mean(K)  ^ 1* 
         mean(B)  ^ 1,
       col = col.alpha("darkblue", 0.7))
text(0.5, 2,"with knowledge&body")
lines (AGE, (1-exp(-1 * AGE  )) ^ 1 * 
         mean(K)  ^ -1* 
         mean(B)  ^ -1,
       col = col.alpha("darkblue", 0.7))
text(1.5, 0.2,"with negative elasticity of knowledge&body")

```

Moving on to $\psi$, this parameter includes all info relative to the foraging trip or trap, specifically, for now, a time scaling parameter, $\xi$ which moderates the effect of the duration of a foraging trip or of how many hours passed since a trap was checked last. 

$$
\psi = time_t \, ^ {\xi}\\
$$
The last parameter in the equations defining $p$ and $\mu$ is $\alpha$, which simply scales all other parameters so that they fit to the data. 
Putting together all the pieces, the values for the probability of success $p$ and the mean amount of returns $\mu$ are shown below. Notice that the two dark curves are made with all the same parameters, but changing the value of $\alpha$, wich indeed scales the whole curve.

```{r put together}
par (mfrow =c(1,2))
plot(NULL, xlim = c(0,3), ylim = c (0,1), 
     xlab = "age", ylab = "returns amount")

for(i in 1:nsamp){
  phi[i,] <- (1-exp(-beta_a[i] * AGE  )) ^ gamma_a[i] *
     mean(K) ^ beta_k[i]
  psi[i] <-    L[2]^lambda[i]
  p <- 1 - exp ( - alpha[i] * phi[i,] * psi[i]) 
  lines( AGE,  p, col = col.alpha("cornflowerblue", 0.3))
}
  phi_1 <- (1-exp(-1 * AGE  )) ^ 1 *
     mean(K) ^ 1
  psi_1 <-    L[3]^1
  p <- 1 - exp ( - 0.2 * phi_1 * psi_1) 
  lines( AGE,  p, col = col.alpha("darkblue", 0.7))

plot(NULL, xlim = c(0,3), ylim = c (0,10), 
     xlab = "age", ylab = "returns amount")

for(i in 1:nsamp){
  phi[i,] <- (1-exp(-beta_a[i] * AGE  )) ^ gamma_a[i] *
     mean(K) ^ beta_k[i]
  psi[i] <-    L[2]^lambda[i]
  R <- exp (log( alpha[i] * phi[i,] * psi[i]) + ((sigma[i]^2) /2))
  lines( AGE,  R, col = col.alpha("cornflowerblue", 0.3))
}
  phi_1 <- (1-exp(-1 * AGE  )) ^ 1 *
     mean(K) ^ 1
  psi_1 <-    L[3]^1
  R <- exp (log( 1 * phi_1 * psi_1) + ((1^2) /2))
  lines( AGE,  R, col = col.alpha("darkblue", 0.7))

```

# APPLY TO REAL DATA

```{r prepare real data}
dc_shellppl <- real_data$shell_ppl[complete.cases(real_data$shell_ppl$knowledge),]
dc_shellppl <- dc_shellppl[complete.cases(dc_shellppl$height),]
dc_shells <- real_data$shells[which(real_data$shells$anonymeID %in% dc_shellppl$anonymeID),]
dat_shells <- list(
  N = nrow(dc_shellppl),
  M = nrow(dc_shells),
  A = dc_shellppl$age[order(dc_shellppl$anonymeID)] / mean(dc_shellppl$age),
  K = dc_shellppl$knowledge[order(dc_shellppl$anonymeID)] / mean(dc_shellppl$knowledge),
  B = dc_shellppl$height[order(dc_shellppl$anonymeID)] / mean(dc_shellppl$height),
  R = as.numeric(dc_shells$returns)/1000,
  L = dc_shells$lenght_min/mean(dc_shells$lenght_min),
  ID_i= as.integer(as.factor(as.character(dc_shells$anonymeID)))
)

dc_trapppl <- real_data$trap_ppl[complete.cases(real_data$trap_ppl$knowledge),]
dc_traps <- real_data$traps[which(real_data$traps$anonymeID %in% dc_trapppl$anonymeID),]
dat_traps <- list(
  N = nrow(dc_trapppl),
  M = nrow(dc_traps),
  A = dc_trapppl$age[order(dc_trapppl$anonymeID)] / mean(dc_trapppl$age),
  K = dc_trapppl$knowledge[order(dc_trapppl$anonymeID)] / mean(dc_trapppl$knowledge),
  B = dc_trapppl$height[order(dc_trapppl$anonymeID)] / mean(dc_trapppl$height),
  S = as.numeric(dc_traps$success),
  L = dc_traps$lenght_hour/mean(dc_traps$lenght_hour),
  ID_i= as.integer(as.factor(as.character(dc_traps$anonymeID)))
)
```

```{r run all models, cache=TRUE}
#age only
m_r <- cstan( file= "../models/Returns_o.stan" , data=dat_shells , chains=3, cores = 3 )
m_s <- cstan( file= "../models/Success_o.stan" , data=dat_traps , chains=3, cores = 3 )
#age&knowledge
m_rk <- cstan( file= "../models/Returns_k.stan" , data=dat_shells , chains=3, cores = 3 )
m_sk <- cstan( file= "../models/Success_k.stan" , data=dat_traps , chains=3, cores = 3 )
#age&body
m_rb <- cstan( file= "../models/Returns_b.stan" , data=dat_shells , chains=3, cores = 3 )
m_sb <- cstan( file= "../models/Success_b.stan" , data=dat_traps , chains=3, cores = 3 )
#age&all
m_ra <- cstan( file= "../models/Returns_all.stan" , data=dat_shells , chains=3, cores = 3 )
m_sa <- cstan( file= "../models/Success_all.stan" , data=dat_traps , chains=3, cores = 3 )

models <- list( shells = c( m_r, m_rk, m_rb, m_ra),
                 traps = c( m_s, m_sk, m_sb, m_sa))
```

First, the model is sampling fine and finding new results.
TO DO: try and move priors for body
```{r plot precis}
titles <- c("age only", "age&knowledge", "age&height", "all")

par(mfrow = c(4,2), mgp = c(1.5, 0.5, 0), mar = c(2.5, 2.5, 2, 1) + 0.1)
for (i in 1:4){
  plot(precis(models$shells[[i]]), main = titles[i] )
  plot(precis(models$traps[[i]]), main = titles[i])
}
```

Then it's also fitting reasonably nicely to the data.
```{r check fit to data}
titles <- c("age only", "age&knowledge", "age&height", "all")
par(mfrow = c(4,2), mgp = c(1.5, 0.5, 0), mar = c(2.5, 2.5, 2, 1) + 0.1)
for (j in 1:4){
  post_r <- extract.samples(models$shells[[j]])
  post_s <- extract.samples(models$traps[[j]])
  plot(dat_shells$A[dat_shells$ID_i] * mean(dc_shellppl$age), dat_shells$R, 
       xlab = "Age", ylab = "Kg shells", main = titles[j],
       xlim = c(0,30), pch = 16, col = col.alpha("grey40", 0.7))
  for(i in 1:150){
  phi <-  exp(apply(post_r$id_v,1,mean )[i]) * (
     (1-exp(- post_r$beta_a[i] * seq(0,3,0.1)  )) ^ post_r$gamma_a[i] *
     ifelse (length(post_r$beta_k) > 0, (mean(dat_shells$K) ) ^ post_r$gamma_k[i] , 1 ) *
     ifelse (length(post_r$beta_b) > 0, (mean(dat_shells$B) ) ^ post_r$gamma_b[i] , 1 )
    )
  psi <-   (mean(dat_shells$L)) ^ post_r$lambda[i]
  R <- exp (  log(post_r$alpha[i] * phi * psi) + (post_r$sigma[i]^2 /2))
  samp_data <- rlnorm(length(seq(0,3,0.1)),  log(post_r$alpha[i] * phi * psi), post_r$sigma[i])
  points(jitter(seq(0,3,0.1)) * mean(dc_shellppl$age), samp_data, col = col.alpha("orange", 0.1), pch = 16)
  lines( seq(0,3,0.1) * mean(dc_shellppl$age),  R, col = col.alpha(shellcol, 0.1), lwd = 2)
  }
  
  plot(jitter(dat_traps$A[dat_traps$ID_i]) * mean(dc_trapppl$age), jitter(dat_traps$S, factor = 0.1), 
       xlab = "Age", ylab = "Prob trap success", main = titles[j],
       xlim = c(0,40), pch = 16, col = col.alpha("grey40", 0.2))
  for(i in 1:150){
  phi <-  exp(apply(post_s$id_v,1,mean )[i]) * (
     (1-exp(- post_s$beta_a[i] * seq(0,3,0.1) ) ) ^ post_s$gamma_a[i] *
     ifelse (length(post_s$beta_k) > 0, (mean(dat_traps$K) ) ^ post_s$gamma_k[i] , 1 ) *
     ifelse (length(post_s$beta_b) > 0, (mean(dat_traps$B) ) ^ post_s$gamma_b[i] , 1 )
    )
  psi <-   mean(dat_traps$L) ^ post_s$lambda[i]
  p <- 1 - exp  ( - post_s$alpha[i] * phi * psi )
  samp_data <- rbern(length(seq(0,3,0.1)), 1 - exp  ( -post_s$alpha[i]  * phi * psi))
  points(jitter(seq(0,3,0.1)) * mean(dc_trapppl$age), samp_data, col = col.alpha("lightgreen", 0.1), pch = 16)
  lines( seq(0,3,0.1) * mean(dc_trapppl$age),  p, col = col.alpha(trapcol, 0.1), lwd = 2)
  }
}
```
